Undecidability and the structure of the Turing degrees

### 2018-08-12 paper.tex
Expanding outline for Turing degree section

\theoremstyle{definition}
\newtheorem{defn}[thm]{Definition}
\newtheorem{defns}[thm]{Definitions}
\newtheorem{con}[thm]{Construction}
\newtheorem{exmp}[thm]{Example}
\newtheorem{exmps}[thm]{Examples}
\newtheorem{notn}[thm]{Notation}
\newtheorem{notns}[thm]{Notations}
\newtheorem{addm}[thm]{Addendum}
\newtheorem{exer}[thm]{Exercise}

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  \cite{turing37_comput_number_with_applic_to_entsc}
  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\subsection{Reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Given two decision problems $A$ and $B$, we say that $A$ is Turing reducible
  to $B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

\begin{prob}[Universality]
  \label{prob:cfg:universality}
  Let $G$ be some context-free grammar over an alphabet $\Sigma$. Then, is
  \[
    L(G) = \Sigma^*?
  \]
  \begin{proof}[Proof of Undecidability]

@@

\section{Turing Degrees}
\begin{defn}
  A Turing degree...
\cite{post44:_recur}
\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{defn}

\begin{defn}
  The jump operator...

\end{defn}

@@

\subsection{Properties and Structure}
\begin{prob}[Post's Problem]
  \cite{post44:_recur}
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
  \cite{Friedberg236}
\end{prob}

@@

>>>>

\theoremstyle{definition}
\newtheorem{defn}[thm]{Definition}
\newtheorem{defns}[thm]{Definitions}
\newtheorem{con}[thm]{Construction}
\newtheorem{exmp}[thm]{Example}
\newtheorem{exmps}[thm]{Examples}
\newtheorem{notn}[thm]{Notation}
\newtheorem{notns}[thm]{Notations}
\newtheorem{addm}[thm]{Addendum}
\newtheorem{exer}[thm]{Exercise}
\newtheorem{fact}[thm]{Fact}

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  \cite{turing37_comput_number_with_applic_to_entsc}
  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

\begin{prob}[Universality]
  \label{prob:cfg:universality}
  Let $G$ be some context-free grammar over an alphabet $\Sigma$. Then, is
  \[
    L(G) = \Sigma^*?
  \]
  \begin{proof}[Proof of Undecidability]\todo{}

@@

\section{Turing Degrees}
In Subsection~\ref{subsec:reducibility}, we introduced the concept of
Turing reducibility, denoted by the symbol $\leq_T$. This symbol suggests some type of
ordering over the set of decision problems. In this section, we introduce that
partial ordering on the set of decision problems and examine its properties.

\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ and $B$ are
  \emph{mutually reducibile} and write $=_T$ if $A \leq_T B$ and $B \leq_T A$.
\end{defn}

\begin{fact}
  The relation $=_T$ is an equivalence relation.
\end{fact}

We call the equivalence classes produced by this relation \emph{Turing degrees}
or \emph{degrees of unsolvability}, a concept first introduced in
1944 by Emil Post~\cite{post44:_recur}. We write the set of all such degrees as $\mathcal{D}$.

\begin{lem}
  The relation $\leq_T$ is a partial ordering of $\mathcal{D}$.
\end{lem}

@@

\subsection{Properties and Structure}
\begin{lem}
  The set $\mathcal{D}$ is of cardinality $2^{\aleph_0}$.
\end{lem}

\begin{lem}
  Every Turing degree contains countably infinite elements.
\end{lem}

While it is simple to explicitly state problems lying in the lower degrees,
in order to more easily study the general structure of $\mathcal{D}$, we
introduce a new operator, \emph{jump}, that increments Turing degrees. Much of
the behavior of this operator was shown in a join paper by Kleene and Post~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}.

\begin{defn}
  The jump operator
\end{defn}

\begin{prob}[Post's Problem]
  \cite{post44:_recur}
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
  \cite{Friedberg236}
\end{prob}

@@

Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}
